A Mechanised Library for Generic (Co)Programming

### 2024-03-05 paper/paper.tex
On Coalgebras

\hypersetup{
  linktocpage,
  colorlinks,
  citecolor=BlueViolet,
  filecolor=red,
  linkcolor=Blue,
  urlcolor=BrickRed
}
\newcommand{\mpav}[1]{\textcolor{red}{\textsc{Marco}: #1}}
\newcommand{\dcas}[1]{\textcolor{ForestGreen}{\textsc{David}: #1}}
\newcommand{\mvol}[1]{\textcolor{blue}{\textsc{Michael}: #1}}
\newcommand{\proofcomment}[1]{\text{\{ #1 \}}}
\newenvironment{proofof}[1] {\begin{proof}[Proof of {#1}]}{\end{proof}}
\newcommand{\eqdef}{\stackrel{\mathrm{\Delta}}{=}}
\newcommand{\bnfeq}{\mathrel{::=}}
\newcommand{\defeq}{\triangleq}
\newcommand{\rul}[3]{\frac{#2}{#3}\;  {\textrulelabel{#1}}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\jud}[3]{#1 \vdash #2 : #3}
\newcommand{\bden}[1]{\llparenthesis#1 \rrparenthesis}
\newcommand{\bigslant}[2]{{\raisebox{.2em}{$#1$}\left/\raisebox{-.2em}{$#2$}\right.}}
\newcommand{\quotient}[2]{\bigslant{#1}{#2}}
\newcommand{\curry}{\Lambda}
\newcommand{\uncurry}{\begin{sideways}\begin{sideways}$\Lambda$\end{sideways}\end{sideways}}
\newcommand{\Bool}{\mathbb{B}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Nat}{\N}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Sets}{\mathbf{Set}}
\newcommand{\blacklater}{\blacktriangleright}
\newcommand{\tot}{\mathcal{S}}
\newcommand{\PSh}{\ensuremath{\textbf{PSh}(\omega)}}

@@

\newcommand{\CatC}{\mathcal{C}}
\newcommand{\CatD}{\mathcal{D}}
\newcommand{\CatE}{\mathcal{E}}
\newcommand{\CatI}{\mathcal{I}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\iso}{\cong}
\newcommand{\ceiling}[1]{\lceil #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}

%\newcommand{\coq}[1]{\mintinline{coq}{#1}}
\newmintinline[coq]{coq}{fontsize=\small}
\newmintinline[ocaml]{OCaml}{fontsize=\small}
\newminted{coq}{fontsize=\small}

@@

\section{Algebras, Coalgebras and Hylomorphisms}
Recursion schemes provide an abstract way to consume and generate data. We know
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise recursive coalgebras (Section~\ref{sec:rec-coalgebras}) and then we
put together these notions to formalise hylomorphisms (Section~\ref{sec:hylo}).

\subsection{Algebras and Catamorphisms}
\label{sec:algebras} An algebra is a set \coq{X} together with some operations
on it\footnote[1]{Sometimes and algebra comes equipped with an equational
theory, e.g. the theory of monoids, but here we focus on algerbraically free
algebras}.  For example, give such a type \coq{X} we can define the monoid
operations as \coq{Empty : unit -> X} and \coq{Concat : X * X -> X}. Notice
that \coq{unit + X * X} is a functor on \coq{X} which means we can equivalently
describe this algebra as a map \coq{unit + X * X -> X} (together with the
monoidal laws).  In this work we will be using algebras to describe the
operations of a data type and so we will not be needing additional equations on
these operations.

@@

Now, the \emph{easiest} way to turn any type \coq{A} into a monoid is to define
it as the least fixed-point of the functor \coq{unit + A + X * X} on \coq{X}.
This is isomorphic to the type of lists over \coq{A} hence it is the least
fixed-point for the functor \coq{unit + A * X} as well. This is the least
``set'' generated by the constructors \coq{Empty : unit -> List A}
and \coq{Cons : A * List A -> List A}. These constructors give an
algebra \coq{LFix_in : unit + A * List A -> List A} which has and inverse which
we  call \coq{LFix_out}.

@@

%%\dcas{Boring stuff}
%The definition of \coq{LFix} in terms of \coq{App} requires us to define our
%own custom induction scheme for initial algebras:
%\begin{coqcode}
%Lemma LFix_ind [P : LFix -> Prop]
%  : (forall x : App F LFix,
%        (forall e : Pos (shape x), P (cont x e)) ->
%        P (LFix_in x))
%    -> forall l : LFix, P l.
%\end{coqcode}
We define smart constructors for the isomorphism of fixed points
\coq{LFix_out}.
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F
l_out : LFix F ~> App F (LFix F)
\end{coqcode}

@@

\subsection{Coalgebras and Anamorphisms}
Terminal coalgebras are defined analogously to initial algebras (see
Figure~\ref{fig:term_coalg}).
% \begin{figure}
\begin{coqcode}
Context `(F : Cont Sh Po).
CoInductive GFix : Type := GFix_in { GFix_out : App F GFix }.
Definition GFixR : GFix -> GFix -> Prop :=...
Instance GFix_setoid : setoid GFix := ...GFixR...
Definition g_in : App F GFix ~> GFix := ...
Definition g_out : GFix ~> App F GFix := ...
\end{coqcode}
% \caption{Terminal Coalgebras}\label{fig:term_coalg}
% \end{figure}
Their equivalence relation, and anamorphisms, and universal properties are
easy to encode, in an analogous way to catamorphisms and initial algebras:
\begin{coqcode}
  Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...

@@

>>>>

\hypersetup{
  linktocpage,
  colorlinks,
  citecolor=BlueViolet,
  filecolor=red,
  linkcolor=Blue,
  urlcolor=BrickRed
}
\newcommand{\mpav}[1]{\textcolor{red}{\textsc{Marco}: #1}}
\newcommand{\dcas}[1]{\textcolor{ForestGreen}{\textsc{David}: #1}}
\newcommand{\mvol}[1]{\textcolor{blue}{\textsc{Michael}: #1}}
\newcommand{\proofcomment}[1]{\text{\{ #1 \}}}
\newenvironment{proofof}[1] {\begin{proof}[Proof of {#1}]}{\end{proof}}
\newcommand{\eqdef}{\stackrel{\mathrm{\Delta}}{=}}
\newcommand{\eqiff}{\stackrel{\triangle}{\iff}}
\newcommand{\bnfeq}{\mathrel{::=}}
\newcommand{\defeq}{\triangleq}
\newcommand{\rul}[3]{\frac{#2}{#3}\;  {\textrulelabel{#1}}}
\newcommand{\den}[1]{\llbracket #1 \rrbracket}
\newcommand{\jud}[3]{#1 \vdash #2 : #3}
\newcommand{\bden}[1]{\llparenthesis#1 \rrparenthesis}
\newcommand{\bigslant}[2]{{\raisebox{.2em}{$#1$}\left/\raisebox{-.2em}{$#2$}\right.}}
\newcommand{\quotient}[2]{\bigslant{#1}{#2}}
\newcommand{\curry}{\Lambda}
\newcommand{\uncurry}{\begin{sideways}\begin{sideways}$\Lambda$\end{sideways}\end{sideways}}
\newcommand{\Bool}{\mathbb{B}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Nat}{\N}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Sets}{\mathbf{Set}}
\newcommand{\blacklater}{\blacktriangleright}
\newcommand{\tot}{\mathcal{S}}
\newcommand{\PSh}{\ensuremath{\textbf{PSh}(\omega)}}

@@

\newcommand{\CatC}{\mathcal{C}}
\newcommand{\CatD}{\mathcal{D}}
\newcommand{\CatE}{\mathcal{E}}
\newcommand{\CatI}{\mathcal{I}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\iso}{\cong}
\newcommand{\ceiling}[1]{\lceil #1 \rceil}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\pair}[2]{\langle #1, #2 \rangle}
\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}
%\newcommand{\coq}[1]{\mintinline{coq}{#1}}
\newmintinline[coq]{coq}{fontsize=\small}
\newmintinline[ocaml]{OCaml}{fontsize=\small}
\newminted{coq}{fontsize=\small}

@@

\section{Formalising Recursion Schemes}
Recursion schemes provide an abstract way to consume and generate data. We know
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise recursive coalgebras (Section~\ref{sec:rec-coalgebras}) and then we
put together these notions to formalise hylomorphisms (Section~\ref{sec:hylo}).

\subsection{Algebras and Catamorphisms}
\label{sec:algebras} An algebra is a set \coq{X} together with some operations
on it\footnote[1]{Sometimes and algebra comes equipped with an equational
theory, e.g. the theory of monoids, but here we focus on algerbraically free
algebras}.  For example, give such a type \coq{X} we can define the monoid
operations as \coq{u : unit -> X} for unit and \coq{m : X * X -> X} for
multiplication. Notice that \coq{FX = unit + X * X} is a functor on \coq{X}
which means we can equivalently describe these two maps as a single one of
type \coq{F X -> X} which we call an algebra for the functor \coq{F}
satisfying the unit and associative laws of the monoid.  In this work we will be
using algebras to describe the operations of a data type and so we will not be
needing additional equations on these operations.

@@

Now, the \emph{easiest} and most effective way to turn any type $A$ into a
monoid is to define it as the least solution to the equation
\begin{coqcode}
Free A = A + unit + Free A * Free A
\end{coqcode}
with \coq{Free : Type -> Type}. In Coq, this is the least fixed-point of the
functor \coq{unit + A + X * X} on \coq{X}.  This is isomorphic to the type of
lists over \coq{A} hence it is the least fixed-point for the
functor \coq{unit + A * X} as well. This is the least ``set'' generated by the
constructors \coq{Empty : unit -> List A} and \coq{Cons : A * List A -> List A}.
These constructors give an algebra \coq{LFix_in : unit + A * List A -> List A}
which has and inverse which we call \coq{LFix_out}.

@@

%%\dcas{Boring stuff}
%The definition of \coq{LFix} in terms of \coq{App} requires us to define our
%own custom induction scheme for initial algebras:
%\begin{coqcode}
%Lemma LFix_ind [P : LFix -> Prop]
%  : (forall x : App F LFix,
%        (forall e : Pos (shape x), P (cont x e)) ->
%        P (LFix_in x))
%    -> forall l : LFix, P l.
%\end{coqcode}
We define smart constructors for the isomorphism of least fixed points
\coq{LFix_out}.
\begin{coqcode}
l_in : App F (LFix F) ~> LFix F
l_out : LFix F ~> App F (LFix F)
\end{coqcode}

@@

\subsection{Coalgebras and Anamorphisms}
While folds \emph{destruct} data, their dual, the unfolds, \emph{construct}
data.  In particular, given a seed value and an \emph{observation}, the unfold
corecursively  ``runs'' the observations starting from the initial seed.

% Consider the transition system below composed by three states \coq{x1}, \coq{x2}
% and \coq{x3} with the transitions labelled by the set of labels \coq{\{a, b\}} and define the map \coq{c} such that it implements the transition system below
% \[\begin{tikzcd}
%     {x_1} & {x_2} & {x_3} \arrow["b"', loop, distance=2em, in=35, out=325]
% 	\arrow["a", from=1-1, to=1-2]
% 	\arrow["b", from=1-2, to=1-3]
% \end{tikzcd}\]

For example, for a type of states \coq{X} and an alphabet type \coq{L} we can
define a (labelled) transition system (LTS) on \coq{X} as a
function \coq{c : X -> L * X} implementing the \emph{transition} map. In
particular, for a state \coq{x1 : X}, \coq{c x1} returns a
pair \coq{(l, x2) : L * X} where \coq{l : L} is the observable action
and \coq{x2 : X} is the next state. We could then add the following notation
\mpav{is this notation duable in Cock?}
\[
  x_{1} \xrightarrow{l} x_{2} \eqiff c x_{1} = (l, x_{2})
\]
\begin{coqcode}
Notation x1 --l> x2 := c(x1) = x2.
\end{coqcode}

Terminal coalgebras are defined analogously to initial algebras (see
Figure~\ref{fig:term_coalg}).

\begin{coqcode}
CoInductive GFix `(F : Cont Sh Po) : Type := GFix_in { GFix_out : App F GFix }.
\end{coqcode}

We define smart constructors for the isomorphism of greatest fixed points
\begin{coqcode}
g_in : App F GFix ~> GFix
g_out : GFix ~> App F GFix
\end{coqcode}
% \caption{Terminal Coalgebras}\label{fig:term_coalg}
% \end{figure}

Their equivalence relation, and anamorphisms, and universal properties are
easy to encode, in an analogous way to catamorphisms and initial algebras:
\begin{coqcode}
  Definition ana `{eA : setoid A} : Coalg F A ~> A ~> GFix...

@@

\begin{coqcode}
Definition GFixR : GFix -> GFix -> Prop :=...
Instance GFix_setoid : setoid GFix := ...GFixR...
\end{coqcode}
